---
title: "TNF_functional_analysis"
author: "Elsie Jacobson"
date: "8/24/2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(Vennerable)
library(extrafont)
library(topGO)
library(org.Hs.eg.db)
library(GSEABase)
library(clusterProfiler)

ggformat <- theme_classic(base_family = "Arial") + theme(text = element_text(size = 20))
```

```{r}
all <- read.csv("../results/HL60_ATRA_TPA_TNF_all.csv")
long <- read.csv("../results/TNF_diff_longform.csv")

TNF1 <- read.csv("../results/TNF_response_level1.csv")
TNF2 <- read.csv("../results/TNF_response_level2.csv")
```

```{r}
tnf_genes <- read.csv("../results/gene_universe_tnf.csv")
g2o_full <- readRDS('../data/biomart_gene_ontology.rds')
g2o <- g2o_full %>%
  filter(go_id != "") %>%
  filter(ensembl_gene_id %in% tnf_genes$x)
geneID2GO <- by(g2o$go_id,
                g2o$ensembl_gene_id,
                function(x) as.character(x))
geneUniverse <- names(geneID2GO)
```

In order to use gene sets in clusterProfiler you need to get them into the clusterProfiler package. There may be a simpler way to do this, but this is how I did it:

Download the hallmark dataset from MSigDB. Select 'entrez gene ids' of the hallmark gene sets
http://software.broadinstitute.org/gsea/msigdb/collections.jsp
Find where the clusterProfiler package is stored with `find.package("clusterProfiler")`
eg. mine is "/Library/Frameworks/R.framework/Versions/3.5/Resources/library/clusterProfiler"
Move the .gmt file to the extdata folder in clusterProfiler
eg. "/Library/Frameworks/R.framework/Versions/3.5/Resources/library/clusterProfiler/extdata/"

```{r}
gmtfile <- system.file("extdata", "h.all.v6.1.entrez.gmt", package="clusterProfiler")
hallmark <- read.gmt(gmtfile)
```

###Gene ontology function

```{r}
find_GO <- function(genes, out, ontology="BP"){
  
  go=paste("../results/functional/GO/",out,"_GO.csv", sep="")
  
  GO_genes <- na.omit(genes[,'ens_gene_p'])
  Interest <- as.character(GO_genes)
  List <- factor(as.integer(geneUniverse %in% GO_genes))
  names(List) <- geneUniverse

  GO <- new("topGOdata", description="", ontology=ontology, allGenes=List,  annot = annFUN.gene2GO, gene2GO = geneID2GO)

  allGO = usedGO(object = GO)

  Fisher.w01 <- runTest(GO, algorithm="weight01", statistic="fisher")

  Res <- GenTable(GO, weight01Fisher = Fisher.w01, orderBy = "weight01Fisher", ranksOf = "weight01Fisher", topNodes = length(allGO))

  sigRes <- Res %>%
  mutate(weight01Fisher=as.numeric(weight01Fisher)) %>%
  filter(weight01Fisher<1) %>%
  filter(Significant>1) %>%
  mutate(qval=p.adjust(weight01Fisher, method="BH")) %>%
  filter(qval<0.05)
  write.csv(sigRes, go)
}

```

ClusterProfiler function

```{r}
cp <- function(genes, out, samples=c("pro+TNF","gran+TNF")){
  
  outkegg=paste("../results/functional/kegg/",out,"_kegg.csv",sep="")
  outenrich=paste("../results/functional/enrichr/",out,"_hall.csv",sep="")
  outenrichgenes=paste("../results/functional/enrichr_genes/",out,"_hall_genes.csv",sep="")
  plothall=paste("../results/functional/plot_enrich/",out,"_hall.pdf",sep="")
  plothallgenes=paste("../results/functional/plot_enrich_genes/",out,"_hall_genes.pdf",sep="")
  
  #get entrez
  entrez <- bitr(genes[,'ens_gene_p'], fromType = "ENSEMBL",
        toType = "ENTREZID",
        OrgDb = org.Hs.eg.db)

  #get kegg
  kegg <- enrichKEGG(gene = entrez$ENTREZID,
                 organism     = 'hsa',
                 pvalueCutoff = 0.05)
  kegg_sig <- kegg@result[kegg@result$p.adjust<0.05,]
  write.csv(kegg_sig,outkegg)
  
  #get hallmark
  
  hall <- enricher(entrez$ENTREZID, TERM2GENE=hallmark)
  hall_plot <- hall[,c('ID','GeneRatio','BgRatio','p.adjust','geneID','Count')]
  write.csv(hall_plot,outenrich)
  
  hall_plot <- hall_plot %>%
    mutate(id=gsub("HALLMARK_","",ID)) %>%
    separate(BgRatio, by='/', into=c('class','total')) %>%
    mutate(Group=as.integer(class)) %>%
    mutate(ratio=Count/Group)
  ggplot(hall_plot, aes(id, ratio, colour=p.adjust)) + geom_point(size=3) + theme_bw(base_family = "Arial") + theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5))
ggsave(plothall,useDingbats=FALSE)

  #plot genes in hallmark categories

  hall_genes <- entrez %>%
    left_join(hallmark, by=c("ENTREZID"="gene")) %>%
    filter(ont %in% hall_plot$ID)

  hall_annot <- genes %>%
    inner_join(hall_genes, by=c("ens_gene_p"="ENSEMBL")) %>%
    dplyr::select(external_gene_name_p,ont)

  hall_annot_tidy <- long %>%
    inner_join(hall_annot, by=c("external_gene_name"="external_gene_name_p")) %>%
    filter(sample %in% samples, sig=="sig", is.na(ont)==FALSE) %>%
    dplyr::select(external_gene_name,sample,sig,log2FoldChange,ens_gene,ont) %>%
    na.omit() %>%
    distinct() %>%
    arrange(ont)

  write.csv(hall_annot_tidy, outenrichgenes)

  ggplot(hall_annot_tidy, aes(external_gene_name, log2FoldChange, fill=sample)) +       geom_bar(stat='identity', position='dodge') + scale_fill_viridis_d() + facet_wrap(~ont, scales = "free_x") + theme_bw() + theme(axis.text.x=element_text(angle=90,hjust=1,vjust=0.5)) 

  ggsave(plothallgenes, height=25, width=25)
}
```

```{r}
go_cluster <- function(genes, out, samples=c("pro+TNF","gran+TNF")){
  find_GO(genes, out)
  cp(genes, out, samples)
}
```

##Pro+TNF vs Gran+TNF

Conserved - same direction

```{r}
pro_gran_same <- na.omit(all[all$padj_p<0.05 & all$padj_g<0.05 & ((all$log2FoldChange_p>0 & all$log2FoldChange_g>0) | (all$log2FoldChange_p<0 & all$log2FoldChange_g<0)),])

go_cluster(pro_gran_same, "pro_gran_conserved")
```

Opposite direction

```{r}
pro_gran_opp <- na.omit(all[all$padj_g<0.05 & all$padj_p<0.05 & ((all$log2FoldChange_g<0 & all$log2FoldChange_p>0) | (all$log2FoldChange_p<0 & all$log2FoldChange_g>0)),])

go_cluster(pro_gran_opp,"pro_gran_opp")
```

Pro not Gran

```{r}
pro_not_gran <- na.omit(all[!(all$padj_g<0.05) & all$padj_p<0.05,])

go_cluster(pro_not_gran,"pro_not_gran")
```

Gran not Pro

```{r}
gran_not_pro <- na.omit(all[all$padj_g<0.05 & all$padj_p>=0.05,])

go_cluster(gran_not_pro,"gran_not_pro")
```

##Include welch diff data

Differentiation conserved - same direction

```{r}
diff_same <- all[all$padj_r<0.05 & all$padj_t<0.05 & ((all$log2FoldChange_r>0 & all$log2FoldChange_t>0) | (all$log2FoldChange_r<0 & all$log2FoldChange_t<0)),]

go_cluster(diff_same, "diff_conserved")
```

```{r}
go_cluster(TNF1,"TNF1")
```

```{r}
go_cluster(TNF2,"TNF2")
```